---
title: "anomaly comparison graphs"
author: "Madhumita Kabade"
date: "2026-04-26"
output: html_document
---

# Description:
#This file is used to prepare and analyse environmental water-quality data
#for anomaly comparison graphs across the three study regions: Bowen, Mackay, and Weipa. The main aim is to build a clean merged dataset containing key physicochemical variables such as temperature, dissolved oxygen, pH, Secchi, salinity, total nitrogen (TN), total phosphorus (TP), and chlorophyll-a together with date, site, region, and seasonal information.

#The workflow in this file will focus on:
#1. cleaning and organising field and lab-based water-quality datasets
#2. creating time-related columns such as year, month, and wet/dry season
#3. summarising environmental variables at the chosen temporal and regional scale
#4. calculating anomalies relative to long-term or study-period means
#5. producing region-wise anomaly comparison graphs to examine how environmental conditions vary through time. 

#These outputs will be used to compare environmental patterns across regions and to support later interpretation of plankton richness, abundance, and community changes in relation to water quality.

# Reading the Physiochem Database File

```{r}
library(readxl)
library(dplyr)
library(lubridate)

#Read
excel_sheets("data/Physiochem Database tables_all_rows(20240902_1226).xlsx")

#Run
field_meas_raw <- read_excel(
  "data/Physiochem Database tables_all_rows(20240902_1226).xlsx",
  sheet = "Field_meas(20240902_1226)"
)

#Inspect
glimpse(field_meas_raw)
unique(field_meas_raw$Site_code)
```

# Filtering the Timeline

```{r}
field_meas_2018_2020 <- field_meas_raw %>%
  filter(Date >= as.Date("2018-01-01"),
         Date <= as.Date("2020-06-30"))

range(field_meas_2018_2020$Date)
nrow(field_meas_2018_2020)

unique(format(field_meas_2018_2020$Date, "%Y"))

anomaly_trial <- field_meas_raw %>%
  filter(Date >= as.Date("2018-01-01"),
         Date <= as.Date("2020-06-30"))
```

# Removing extra sites and adding a region column 


```{r}
sites_keep <- c(
  # Bowen
  "AP_AMB1", "AP_AMB2", "AP_AMB3", "AP_AMB4", "AP_AMB5",

  # Mackay
  "MKY_AMB1", "MKY_AMB2", "MKY_AMB3B", "MKY_AMB5", "MKY_AMB6B",
  "MKY_AMB8", "MKY_AMB10", "MKY_AMB11", "MKY_AMB12",
  "MKY_CAM1", "MKY_CAM2", "MKY_CAM3",

  # Weipa
  "WP_AMB1", "WP_AMB2", "WP_AMB3", "WP_AMB4", "WP_AMB5"
)

anomaly_trial <- field_meas_raw %>%
  filter(Date >= as.Date("2018-01-01"),
         Date <= as.Date("2020-06-30")) %>%
  filter(Site_code %in% sites_keep)

# check if its working

print(unique(anomaly_trial$Site_code))
print(nrow(anomaly_trial))

# See how many rows per site
print(
  anomaly_trial %>%
    count(Site_code, sort = TRUE)
)

# adding region columns

anomaly_trial <- anomaly_trial %>%
  mutate(
    region = case_when(
      grepl("^AP_", Site_code) ~ "Bowen",
      grepl("^MKY_", Site_code) ~ "Mackay",
      grepl("^WP_", Site_code) ~ "Weipa",
      TRUE ~ NA_character_
    )
  )

# check if regions are properly mapped

print(
  anomaly_trial %>%
    distinct(Site_code, region) %>%
    arrange(region, Site_code)
)
```

# Testing if Secchi has only one value per site+date combination or multiple

```{r}
field_obs_raw <- read_excel(
  "data/Physiochem Database tables_all_rows(20240902_1226).xlsx",
  sheet = "Field_obs(20240902_1226)"
)

secchi_tbl <- field_obs_raw %>%
  filter(Date >= as.Date("2018-01-01"),
         Date <= as.Date("2020-06-30")) %>%
  filter(Site_code %in% sites_keep) %>%
  mutate(Date = as.Date(Date)) %>%
  select(Site_code, Date, Secchi)

secchi_tbl %>%
  count(Site_code, Date) %>%
  filter(n > 1)

field_obs_raw %>%
  mutate(Date = as.Date(Date)) %>%
  count(Site_code, Date) %>%
  filter(n > 1)

```

# Adding Secchi to the anamoly_trial databse

```{r}
# 1. Read field observations
field_obs_raw <- read_excel(
  "data/Physiochem Database tables_all_rows(20240902_1226).xlsx",
  sheet = "Field_obs(20240902_1226)"
)

# Create secchi_tbl
secchi_tbl <- field_obs_raw %>%
  mutate(Date = as.Date(Date)) %>%
  filter(Date >= as.Date("2018-01-01"),
         Date <= as.Date("2020-06-30")) %>%
  filter(Site_code %in% sites_keep) %>%
  select(Site_code, Date, Secchi)

# Check it looks right
print(secchi_tbl)
print(nrow(secchi_tbl))

# Make sure Date in anomaly_trial is also a plain Date
anomaly_trial <- anomaly_trial %>%
  mutate(Date = as.Date(Date))

# Join Secchi into anomaly_trial
anomaly_trial <- anomaly_trial %>%
  left_join(secchi_tbl, by = c("Site_code", "Date"))

# Check if it worked
print(names(anomaly_trial))
print(sum(is.na(anomaly_trial$Secchi)))

# And to see the data
head(anomaly_trial)

# Check which site-date combinations did not get a Secchi value
print(
  anomaly_trial %>%
    filter(is.na(Secchi)) %>%
    distinct(Site_code, Date) %>%
    arrange(Site_code, Date)
)

cat("n =", nrow(anomaly_trial %>% filter(is.na(Secchi))), "\n")

print(
  anomaly_trial %>%
    filter(is.na(Secchi)) %>%
    arrange(Site_code, Date)
)

#Download to show Nathan
library(writexl)

write_xlsx(anomaly_trial,
           "data/ongoing_trial_anomaly_dataset.xlsx")


# removing extra secchi columns
anomaly_trial <- anomaly_trial %>%
  select(-starts_with("Secchi."))   # removes Secchi.x, Secchi.y, etc.

```