---
title: "plankton richness and abundance time series"
author: "Madhumita Kabade"
date: "2025-08-24"
output: html_document
Description: This project explores temporal patterns in zooplankton and phytoplankton
  communities across multiple coastal sites. The focus is on tracking changes
  in plankton richness (number of taxa) and abundance (total counts) over time,
  and comparing these patterns between regions. By visualising and analysing
  date-by-date time series, the project aims to identify seasonal peaks and lows,
  evaluate spatial differences between sites, and provide a foundation for
  further comparisons with water quality variables such as nutrients, turbidity,
  temperature, and salinity.
---

####loading packages

```{r}
library(readxl)    # to read Excel files
library(dplyr)     # for data wrangling
library(janitor)   # for cleaning column names
library(lubridate) # for handling dates
```

####loading data files

```{r}
# Define file paths (relative to project root)
zoop_file  <- "../data/NQBP zooplankton 2021-02-15.xlsx"
phyto_file <- "../data/NQBP phytoplankton 2021-02-18.xlsx"

# Import sheets
zoop_raw <- read_excel(zoop_file, sheet = "in") |> clean_names()
phyto_raw <- read_excel(phyto_file, sheet = "in") |> clean_names()

# Quick peek
head(zoop_raw)
head(phyto_raw)
```

####setting the scene,prepping for the data frame

```{r}
# Ensure Date type
zoop <- zoop_raw |> dplyr::mutate(sample_date = as.Date(sample_date))
phyto <- phyto_raw |> dplyr::mutate(sample_date = as.Date(sample_date))

# Add region if missing (edit mapping if yours differ)
add_region <- function(df){
  has_region <- "region" %in% names(df)
  if (has_region) return(df)
  df |>
    dplyr::mutate(
      site_code = dplyr::coalesce(site_code, site_name),
      region_code = stringr::str_extract(site_code, "^[A-Za-z]{3}"),
      region = dplyr::case_when(
        region_code %in% c("MKY") ~ "Mackay",
        region_code %in% c("BWN") ~ "Bowen",
        region_code %in% c("WEI","WEP") ~ "Weipa",
        TRUE ~ "Unknown"
      )
    ) |>
    dplyr::select(-region_code)
}
zoop <- add_region(zoop)
phyto <- add_region(phyto)
```

####quick checks

```{r}
# Check types
str(zoop$sample_date); str(phyto$sample_date)

# Peek at the mapping results
# Zooplankton: include site_name
dplyr::count(zoop, site_code, site_name, region, sort = TRUE)

# Phytoplankton: include site_name
dplyr::count(phyto, site_code, site_name, region, sort = TRUE)

# See if anything fell into "Unknown"
dplyr::count(zoop, region)
dplyr::count(phyto, region)

# Earliest/latest sampling dates (helps set plot x-limits)
range(zoop$sample_date, na.rm = TRUE)
range(phyto$sample_date, na.rm = TRUE)

# Any missing dates?
sum(is.na(zoop$sample_date))
sum(is.na(phyto$sample_date))

```

####loading packages for analysis

```{r}
# Data wrangling + plotting
library(dplyr)
library(ggplot2)
library(readr)   # for write_csv()

```

####building summary tables fpr zooplankton and phytoplankton - abundance and richness

```{r}
# Abundance: sum of counts (orgs) per date/site/region
zoop_abund_daily <- zoop |>
  dplyr::group_by(sample_date, site_code, region) |>
  dplyr::summarise(
    zoop_total = sum(as.numeric(orgs), na.rm = TRUE),
    .groups = "drop"
  )

# Richness: number of distinct taxa per date/site/region
zoop_rich_daily <- zoop |>
  dplyr::filter(!is.na(taxa) & taxa != "") |>
  dplyr::group_by(sample_date, site_code, region) |>
  dplyr::summarise(
    zoop_richness = dplyr::n_distinct(taxa),
    .groups = "drop"
  )

# Join abundance + richness; order for tidy viewing
zoop_daily <- dplyr::full_join(
  zoop_abund_daily, zoop_rich_daily,
  by = c("sample_date","site_code","region")
) |>
  dplyr::arrange(site_code, sample_date)    

# ---- PHYTOPLANKTON ----

# Abundance: sum of cell_count per date/site/region
phyto_abund_daily <- phyto |>
  dplyr::group_by(sample_date, site_code, region) |>
  dplyr::summarise(
    phyto_total = sum(as.numeric(cell_count), na.rm = TRUE),
    .groups = "drop"
  )

# Richness: number of distinct taxa per date/site/region
phyto_rich_daily <- phyto |>
  dplyr::filter(!is.na(taxa) & taxa != "") |>
  dplyr::group_by(sample_date, site_code, region) |>
  dplyr::summarise(
    phyto_richness = dplyr::n_distinct(taxa),
    .groups = "drop"
  )

# Join abundance + richness; order for tidy viewing
phyto_daily <- dplyr::full_join(
  phyto_abund_daily, phyto_rich_daily,
  by = c("sample_date","site_code","region")
) |>
  dplyr::arrange(site_code, sample_date)

#saving the output summary tables
readr::write_csv(zoop_daily,  "output/zooplankton_daily_abund_rich.csv")
readr::write_csv(phyto_daily, "output/phytoplankton_daily_abund_rich.csv")


```

####debugging why output file isnt saving

```{r}
# Make sure the main output folder exists
if(!dir.exists("../output")) dir.create("../output")

# Move files from code/output → output
file.rename("output/zooplankton_daily_abund_rich.csv", 
            "../output/zooplankton_daily_abund_rich.csv")

file.rename("output/phytoplankton_daily_abund_rich.csv", 
            "../output/phytoplankton_daily_abund_rich.csv")
```

####checking if any data rows were lost when summary tables were formed

```{r}
# Original row counts
nrow(zoop_raw)
nrow(phyto_raw)

# Summarised row counts (unique date × site × region combos)
nrow(zoop_daily)
nrow(phyto_daily)

# How many unique sampling events were in the raw data?
n_distinct(zoop_raw$sample_date)
n_distinct(phyto_raw$sample_date)

# Total abundance in raw vs. summarised
sum(as.numeric(zoop_raw$orgs), na.rm = TRUE)
sum(zoop_daily$zoop_total, na.rm = TRUE)

sum(as.numeric(phyto_raw$cell_count), na.rm = TRUE)
sum(phyto_daily$phyto_total, na.rm = TRUE)

# recompute richness from raw for every site/date
check_rich <- zoop_raw |>
  group_by(site_code, sample_date) |>
  summarise(raw_rich = n_distinct(taxa), .groups = "drop")

# join with your summary table
compare <- left_join(check_rich, zoop_daily,
                     by = c("site_code","sample_date"))

# any site/date where richness differs?
compare |> filter(raw_rich != zoop_richness)

# site-date combos present in raw but missing in summary
anti_join(check_rich, zoop_daily, by = c("site_code","sample_date"))

## 2) Recompute richness from raw for every site × date
phyto_check_rich <- phyto_raw |>
  group_by(site_code, sample_date) |>
  summarise(raw_rich = n_distinct(taxa), .groups = "drop")

## 3) Compare raw-recomputed richness to your summarised table
phyto_compare <- left_join(
  phyto_check_rich,
  phyto_daily,                     # has phyto_richness
  by = c("site_code", "sample_date")
)

# Any mismatches?
phyto_compare |> filter(raw_rich != phyto_richness)
# Expect: 0 rows

## 4) Coverage: any site × date in raw missing from summary?
anti_join(phyto_check_rich, phyto_daily, by = c("site_code","sample_date"))
# Expect: 0 rows

```

####sorting both the tables from highest count to lowest count

```{r}
library(dplyr)

# ---- Phytoplankton sorted ----
# By abundance (highest first)
phyto_daily_abund_sorted <- phyto_daily %>%
  arrange(desc(phyto_total))

# By richness (highest first)
phyto_daily_rich_sorted <- phyto_daily %>%
  arrange(desc(phyto_richness))

# ---- Zooplankton sorted ----
# By abundance (highest first)
zoop_daily_abund_sorted <- zoop_daily %>%
  arrange(desc(zoop_total))

# By richness (highest first)
zoop_daily_rich_sorted <- zoop_daily %>%
  arrange(desc(zoop_richness))

# Peek at top 10 rows of each
head(phyto_daily_abund_sorted, 10)
head(phyto_daily_rich_sorted, 10)
head(zoop_daily_abund_sorted, 10)
head(zoop_daily_rich_sorted, 10)

# ---- Update phyto_daily ----
phyto_daily <- phyto_daily %>%
  arrange(desc(phyto_total), desc(phyto_richness))

# ---- Update zoop_daily ----
zoop_daily <- zoop_daily %>%
  arrange(desc(zoop_total), desc(zoop_richness))

```

####making region level tables, for region based analysis

```{r}
library(dplyr)
library(ggplot2)
library(lubridate)   # for floor_date / ceiling_date

# ---------- REGION-DAILY TABLES (as you had) ----------
# ZOOP
zoop_region_abund <- zoop %>%
  group_by(sample_date, region) %>%
  summarise(zoop_total = sum(as.numeric(orgs), na.rm = TRUE), .groups = "drop")

zoop_region_rich <- zoop %>%
  filter(!is.na(taxa) & taxa != "") %>%
  group_by(sample_date, region) %>%
  summarise(zoop_richness = n_distinct(taxa), .groups = "drop")

zoop_region_daily <- full_join(
  zoop_region_abund, zoop_region_rich,
  by = c("sample_date","region")
) %>% arrange(region, sample_date)

# PHYTO
phyto_region_abund <- phyto %>%
  group_by(sample_date, region) %>%
  summarise(phyto_total = sum(as.numeric(cell_count), na.rm = TRUE), .groups = "drop")

phyto_region_rich <- phyto %>%
  filter(!is.na(taxa) & taxa != "") %>%
  group_by(sample_date, region) %>%
  summarise(phyto_richness = n_distinct(taxa), .groups = "drop")

phyto_region_daily <- full_join(
  phyto_region_abund, phyto_region_rich,
  by = c("sample_date","region")
) %>% arrange(region, sample_date)

# ---------- CONSISTENT X-AXIS (3-month ticks, same limits for all plots) ----------
x_min <- min(c(zoop_region_daily$sample_date,  phyto_region_daily$sample_date), na.rm = TRUE)
x_max <- max(c(zoop_region_daily$sample_date,  phyto_region_daily$sample_date), na.rm = TRUE)

# snap limits nicely to quarter starts/ends
x_lim_min <- floor_date(x_min, "quarter")
x_lim_max <- ceiling_date(x_max, "quarter")

x_scale <- scale_x_date(
  limits      = c(x_lim_min, x_lim_max),
  breaks      = seq(x_lim_min, x_lim_max, by = "3 months"),  # Jan/Apr/Jul/Oct pattern
  date_labels = "%Y-%m",
  expand      = c(0.01, 0.01)
)

common_theme <- theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# ---------- PLOTS ----------
# 1) Zooplankton abundance (by region)
p1 <- ggplot(zoop_region_daily, aes(sample_date, zoop_total, color = region)) +
  geom_line() + geom_point(size = 0.8) +
  labs(title = "Zooplankton abundance over time (by region)",
       x = "Date", y = "Total zooplankton") +
  x_scale + common_theme

# 2) Zooplankton richness (by region)
p2 <- ggplot(zoop_region_daily, aes(sample_date, zoop_richness, color = region)) +
  geom_line() + geom_point(size = 0.8) +
  labs(title = "Zooplankton richness over time (by region)",
       x = "Date", y = "Number of taxa") +
  x_scale + common_theme

# 3) Phytoplankton abundance (by region)
p3 <- ggplot(phyto_region_daily, aes(sample_date, phyto_total, color = region)) +
  geom_line() + geom_point(size = 0.8) +
  labs(title = "Phytoplankton abundance over time (by region)",
       x = "Date", y = "Total phytoplankton (cells)") +
  x_scale + common_theme

# 4) Phytoplankton richness (by region)
p4 <- ggplot(phyto_region_daily, aes(sample_date, phyto_richness, color = region)) +
  geom_line() + geom_point(size = 0.8) +
  labs(title = "Phytoplankton richness over time (by region)",
       x = "Date", y = "Number of taxa") +
  x_scale + common_theme

# print
p1; p2; p3; p4


```

####code for zooplankton sampling effort

```{r}
# Zooplankton sampling effort (count per region per date)
zoop_sampling <- zoop %>%
  dplyr::count(sample_date, region, name = "n_samples")

# ---- Zooplankton sampling effort ----
ggplot(zoop_sampling, aes(x = sample_date, y = n_samples, color = region)) +
  geom_line() + geom_point(size = 0.8) +
  labs(title = "Zooplankton sampling effort (by region)",
       x = "Date", y = "Number of samples") +
  scale_x_date(
    date_breaks = "3 months",   # every 3 months
    date_labels = "%Y-%m"       # format like 2018-01, 2018-04, etc
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

####code for phytoplankton sampling effort

```{r}
# Phytoplankton sampling effort (count per region per date)
phyto_sampling <- phyto %>%
  dplyr::count(sample_date, region, name = "n_samples")

ggplot(phyto_sampling, aes(x = sample_date, y = n_samples, color = region)) +
  geom_line() + geom_point(size = 0.8) +
  labs(title = "Phytoplankton sampling effort (by region)",
       x = "Date", y = "Number of samples") +
  scale_x_date(
    date_breaks = "3 months",   
    date_labels = "%Y-%m"       
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


####site wise time series comparisons

```{r}
install.packages(c("ggforce","scales"))  # if needed
library(dplyr)
library(ggplot2)
library(ggforce)
library(scales)
library(grid)   # for unit()

plot_sites_ts <- function(df, region_name, ycol, ylab,
                          page = 1, nrow = 3, ncol = 3) {
  df %>%
    filter(region == region_name) %>%
    group_by(site_code) %>%
    mutate(.order = median(.data[[ycol]], na.rm = TRUE)) %>%
    ungroup() %>%
    ggplot(aes(x = sample_date, y = .data[[ycol]])) +
    geom_line(linewidth = 0.5, color = "#2E2E2E") +
    geom_point(size = 1.1, alpha = 0.85, color = "#2E2E2E") +
    scale_x_date(
      date_breaks = "3 months",     # <- unchanged
      date_labels = "%Y-%m",
      expand = expansion(mult = c(0.01, 0.03))
    ) +
    scale_y_continuous(labels = label_number(big.mark = ",")) +
    labs(title = paste(ylab, "by site —", region_name),
         x = "Date", y = ylab) +
    facet_wrap_paginate(
      ~ reorder(site_code, .order),
      nrow = nrow, ncol = ncol, page = page,
      scales = "free_y"
    ) +
    theme_minimal(base_size = 10) +
    theme(
      panel.grid.minor = element_blank(),
      panel.spacing = unit(10, "pt"),
      strip.text = element_text(face = "bold", size = 10),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 8),
      axis.text.y = element_text(size = 8),
      plot.title = element_text(face = "bold", size = 13),
      plot.margin = margin(6, 10, 6, 10)
    )
}


# ===================== ZOOPLANKTON =====================

# Abundance by site, per region
plot_sites(zoop_daily, "Bowen",  "zoop_total",     "Total zooplankton",            "Zooplankton abundance by site", page = 1)
plot_sites(zoop_daily, "Mackay", "zoop_total",     "Total zooplankton",            "Zooplankton abundance by site", page = 2)
plot_sites(zoop_daily, "Weipa",  "zoop_total",     "Total zooplankton",            "Zooplankton abundance by site", page = 1)

# Richness by site, per region
plot_sites(zoop_daily, "Bowen",  "zoop_richness",  "Zooplankton taxa richness",    "Zooplankton richness by site", page = 1)
plot_sites(zoop_daily, "Mackay", "zoop_richness",  "Zooplankton taxa richness",    "Zooplankton richness by site", page = 2)
plot_sites(zoop_daily, "Weipa",  "zoop_richness",  "Zooplankton taxa richness",    "Zooplankton richness by site", page = 1)

# ===================== PHYTOPLANKTON =====================

# Abundance by site, per region
plot_sites(phyto_daily, "Bowen",  "phyto_total",    "Total phytoplankton (cells)",  "Phytoplankton abundance by site", page = 1)
plot_sites(phyto_daily, "Mackay", "phyto_total",    "Total phytoplankton (cells)",  "Phytoplankton abundance by site", page = 2)
plot_sites(phyto_daily, "Weipa",  "phyto_total",    "Total phytoplankton (cells)",  "Phytoplankton abundance by site", page = 1)

# Richness by site, per region
plot_sites(phyto_daily, "Bowen",  "phyto_richness", "Phytoplankton taxa richness",  "Phytoplankton richness by site", page = 1)
plot_sites(phyto_daily, "Mackay", "phyto_richness", "Phytoplankton taxa richness",  "Phytoplankton richness by site", page = 2)
plot_sites(phyto_daily, "Weipa",  "phyto_richness", "Phytoplankton taxa richness",  "Phytoplankton richness by site", page = 1)

```

